---
title: "Fulton County COVID Stats"
author: "Melinda Higgins"
date: "2022-11-11"
output:
  html_document: default
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# set up knitr options
# echo - show/hide code in report
# error - set to TRUE keep knitting even if errors
# error - set to FALSE to stop knitting on errors
# message - FALSE turn off messages
# warning - FALSE turn off warnings

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# load packages
library(dplyr)
```

# Report Overview

This report will help you use all of the skills you've been introduced to during this R workshop including:

* making tables
* making charts and figures
* reading in datafiles
* creating new variables
* merging datasets
* and some new skills - making maps!!

## Fulton County COVID Data

```{r}
# read in the COVID data EXCEL file
library(readxl)
coviddat <- 
  read_excel("FALSE COVID DATA_TRAINING.xlsx")

# get number of cases
n_covid <- dim(coviddat)[1]

# get min and max dates
mindate <- min(coviddat$reprt_creationdt_FALSE)
maxdate <- max(coviddat$reprt_creationdt_FALSE)

# learn about date formatting
# https://epirhandbook.com/en/working-with-dates.html#date-display
```

The dataset you will be working with has `r n_covid` COVID cases from mostly the Fulton County, GA area. These data cover COVID cases reported from `r format(mindate, format = "%B %Y")` to `r format(maxdate, format = "%B %Y")`.

### Table of Demographics

```{r}
library(arsenal)

tab1 <- 
  tableby(~ case_age + case_gender +
            case_race + case_eth, 
          data=coviddat)
```

```{r results="asis"}
summary(tab1)
```

### Table of symptoms

```{r results="asis"}
tab1 <- 
  tableby(~ sym_fever + sym_subjfever + sym_myalgia +
            sym_losstastesmell + sym_sorethroat +
            sym_cough + sym_headache + sym_resolved, 
          data=coviddat)

summary(tab1)
```

## Clean up data and recode

```{r}
# recode race into
# black, white and other
# where other is the other races, unknowns and NAs (missing)
coviddat <-
  coviddat %>%
  mutate(
    case_race.c = case_when(
      case_race == "BLACK" ~ "Black",
      case_race == "WHITE" ~ "White",
      is.na(case_race) ~ "Other, unknown or missing",
      TRUE ~ "Other, unknown or missing"
    )
  )

# also clean up some of the symptoms
# keep yes and compare to no, unknown or NA missing
# recode with case_when

# sym_fever
coviddat <-
  coviddat %>%
  mutate(
    sym_fever.c = case_when(
      sym_fever == "No" ~ "No, unk, na",
      sym_fever == "Unk" ~ "No, unk, na",
      sym_fever == "Yes" ~ "Yes",
      is.na(sym_fever) ~ "No, unk, na",
      TRUE ~ "No, unk, na"
    )
  )

# sym_sorethroat
coviddat <-
  coviddat %>%
  mutate(
    sym_sorethroat.c = case_when(
      sym_sorethroat == "No" ~ "No, unk, na",
      sym_sorethroat == "Unk" ~ "No, unk, na",
      sym_sorethroat == "Yes" ~ "Yes",
      is.na(sym_sorethroat) ~ "No, unk, na",
      TRUE ~ "No, unk, na"
    )
  )

# sym_headache
coviddat <-
  coviddat %>%
  mutate(
    sym_headache.c = case_when(
      sym_headache == "No" ~ "No, unk, na",
      sym_headache == "Unk" ~ "No, unk, na",
      sym_headache == "Yes" ~ "Yes",
      is.na(sym_headache) ~ "No, unk, na",
      TRUE ~ "No, unk, na"
    )
  )

# sym_cough
coviddat <-
  coviddat %>%
  mutate(
    sym_cough.c = case_when(
      sym_cough == "No" ~ "No, unk, na",
      sym_cough == "Unk" ~ "No, unk, na",
      sym_cough == "Yes" ~ "Yes",
      is.na(sym_cough) ~ "No, unk, na",
      TRUE ~ "No, unk, na"
    )
  )
```

### Table of some symptoms after recoding

```{r results="asis"}
tab1 <- 
  tableby(~ sym_fever.c + sym_sorethroat.c +
            sym_cough.c + sym_headache.c, 
          data=coviddat)

summary(tab1)
```

### Table of symptoms by race

Let's also add some nicer labels and then make the table of these recoded symptoms by race.

```{r results="asis"}
attr(coviddat$case_race.c,'label')  <- 'Race'
attr(coviddat$sym_cough.c,'label')  <- 'Cough'
attr(coviddat$sym_fever.c,'label')  <- 'Fever'
attr(coviddat$sym_headache.c,'label')  <- 'Headache'
attr(coviddat$sym_sorethroat.c,'label')  <- 'Sore Throat'

tab1 <- 
  tableby(case_race.c~ sym_fever.c + sym_sorethroat.c +
            sym_cough.c + sym_headache.c, 
          test = FALSE,
          data=coviddat)

summary(tab1,
        title = "Symptoms by race")
```

### Table of Symptoms by Race for People who Died

```{r results = "asis"}
coviddat.died <- 
  coviddat %>%
  filter(died == "Yes")
  
tab1 <- 
  tableby(case_race.c~ sym_fever.c + sym_sorethroat.c +
            sym_cough.c + sym_headache.c, 
          test = FALSE,
          data=coviddat.died)

summary(tab1,
        title = "Symptoms by race - people who died")
```


# Maps with points and markers

Making maps with R is almost a whole other workshop in and of itself. But let's try a few simple maps to get started.

In the exercises below you will be using these packages:

* `leaflet` - learn more at [https://rstudio.github.io/leaflet/](https://rstudio.github.io/leaflet/)
* `sf` - here are places to learn more:
    - [https://r-spatial.github.io/sf/](https://r-spatial.github.io/sf/)
    - [http://www.geo.hunter.cuny.edu/~ssun/R-Spatial/mapping.html#plotting-simple-features-sf-with-plot](http://www.geo.hunter.cuny.edu/~ssun/R-Spatial/mapping.html#plotting-simple-features-sf-with-plot)

## Make a Map of Atanta with `leaflet`

Learn more about the [`leaflet` package](https://rstudio.github.io/leaflet/).

```{r}
library(leaflet)
library(dplyr)

# Fulton County Lat Long
# 33.8034° N, 84.3963° W

m <- leaflet() %>% 
  setView(lng = -84.3963, lat = 33.8034, zoom = 12)

# add default open street map "tile"
m %>% addTiles()
```

## Add data to the map - points/markers

Let's look at the locations of the COVID deaths under review and color the points by gender.

```{r}
df1 <- coviddat %>%
  filter(coviddat$died_covid == "Under Review") %>%
  select(PID, case_gender,
         latitude_JITT, longitude_JITT)

# add some color by gender
df1$colorgender <- 
  ifelse(
    df1$case_gender == "Female", "darkorange", 
    "blue"
  )

# change zoom to 10 to zoom out a little (wider area)
m2 <- leaflet(df1) %>% 
  setView(lng = -84.3963, lat = 33.8034, zoom = 10) 

m2 %>%
  addTiles() %>%
  addCircles(lng = ~longitude_JITT, 
             lat = ~latitude_JITT,
             color = ~colorgender)
```

## Change the base layer if wanted to simplier road map

Learn more with basemaps at [https://rstudio.github.io/leaflet/basemaps.html](https://rstudio.github.io/leaflet/basemaps.html).

```{r}
m %>% addProviderTiles(providers$MtbMap) %>%
  addProviderTiles(
    providers$Stamen.TonerLines,
    options = providerTileOptions(opacity = 0.35)
    ) %>%
  addProviderTiles(providers$Stamen.TonerLabels)
```

## Add some data to the map

```{r}
# change zoom back to 11 to zoom back in a little
m2 <- leaflet(df1) %>% 
  setView(lng = -84.3963, lat = 33.8034, zoom = 11)

m2 %>% addProviderTiles(providers$MtbMap) %>%
  addProviderTiles(
    providers$Stamen.TonerLines,
    options = providerTileOptions(opacity = 0.35)
    ) %>%
  addProviderTiles(providers$Stamen.TonerLabels) %>%
  addCircles(lng = ~longitude_JITT, 
             lat = ~latitude_JITT,
             color = ~colorgender)
```

# Aggregating data

In this part of the report, you will:

* use the [`sf`](https://r-spatial.github.io/sf/index.html) package to lay out the geometry shapefiles for the zipcode boundaries in Fulton County, GA
* use the [`data.table`](https://rdatatable.gitlab.io/data.table/) package to create an aggregated dataset with summaries of variables by zipcode
* use the [`dplyr`](https://dplyr.tidyverse.org/) package to join the aggregated dataset variable to the shapefile dataset - see some examples at [https://statisticsglobe.com/aggregate-data-table-group-r](https://statisticsglobe.com/aggregate-data-table-group-r)
* and use the merged dataset to create a choropleth map

## Load shapefiles for Fulton County Map

Read in the SHP shapefile for Fulton County, GA using the `sf` package. Then plot the map.

```{r}
library(sf)
fulton_sf <- 
  sf::st_read(
    "FultonCountyZipCodes/FultonCountyZipCodes.shp"
    )

plot(st_geometry(fulton_sf), 
     main='Fulton County Geometry')
```

## Pull data of interest and summarize by zipcode

```{r}
library(data.table)

# see cough symptoms - how codes
# what is missing or unknown
# table(coviddat$sym_cough, useNA = "ifany")

# use data.table for aggregation
# do simple counts for now
cough_zip <-
  coviddat %>%
  select(PID, case_zip, sym_cough, sym_cough.c)

cough_zip <- as.data.table(cough_zip)
#class(cough_zip)

cough_zip$sym_cough.cnum <-
  as.numeric(cough_zip$sym_cough.c == "Yes")

# aggregate data using data.table functions
cough_sum <- 
  cough_zip[ , .(cough_sum = sum(sym_cough.c == "Yes")), 
             by = "case_zip"]
```


## Then merge and make a choropleth map

```{r}
# rname case_zip to ZipCode
# and change to character
cough_sum$ZipCode <- as.character(cough_sum$case_zip)

fulton_sf_cough <- left_join(
  fulton_sf, cough_sum, by="ZipCode"
)

# this uses library(sf)
plot(fulton_sf_cough['cough_sum'], 
     main='Number of People with Cough Reported')
```

## Another map option with ggplot2

```{r}
# see colors and options
# at https://r-spatial.github.io/sf/articles/sf5.html

library(ggplot2)
ggplot() + 
  geom_sf(data = fulton_sf_cough, 
          aes(fill = cough_sum)) + 
  scale_y_continuous(n.breaks=20)
```

